Energy Clustering

### 2017-09-30 energy_clustering/draft/energy_clustering.tex
initialization issues

%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{mathtools}
%\usepackage{printlen}
\usepackage[inline]{enumitem}
%\usepackage{cite}
%\usepackage{bbm}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}

\hyphenation{op-tical net-works semi-conduc-tor}

@@

%% our definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\LC}{LC}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\OO{{\mathcal{O}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}
%\newcommand{\Ind}[1]{\delta_{#1}}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

@@

\author{Guilherme Fran\c ca}
\email{guifranca@gmail.com} 
%\affiliation{Johns Hopkins University}

\author{Joshua T. Vogelstein}
\email{jovo@jhu.edu}

\affiliation{Johns Hopkins University}

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
%Although  
%extensively used in statistics, it
%has not been previously incorporated in machine learning problems.
% @gui: not sure what you mean?  what counts as machine learning?  be specific, i think yo mean clustering?  even then, it is not true, both the master's thesis and the RKHS paper kind of do it.  be as general as possible about the gap that we are filling, but not more general.
% @jovo: commented out, agree
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else? 
% @jovo: any kernel method is defined on a HKHS. 
% For instance ref [11] is the first one to consider kernel k-means in a 
% RKHS. The whole problem is how do you formulate a 
% statistical problem in a RKHS, or how do you know if there is a kernel
% related to it. In the case of energy this was done by [6], and it's based 
% on this work that we have the theory presented in this paper
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models in a variety of settings.
\end{abstract}

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution which
is achieved under minimum energy. When probability distributions
are different, this statistical potential 
energy diverges as sample size increases.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, nonlinear
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and  
hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance. We refer to \cite{Szkely2013} and 
references therein for an overview.
Moreover, an energy statistics formulation
to clustering was already proposed in \cite{Kgroups} 
which greatly motivated this paper.

Recently, distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
established \cite{Sejdinovic2013}. This provides the 
link between concepts used in energy statistics and concepts
commonly used in machine learning, and form the basis of our approach.

The clustering problem has a long history in machine learning.
Perhaps the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and it is thus sensitive to outliers. 
% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.
% @jovo: commented out below and included a phrase in the end
%Moreover, $k$-means
%is strongly parametric, assuming that data is isotropically and normally
%distributed since it is a special case of gaussian mixture models (GMM) 
% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.  k-means is certainly closely related to GMM though.
%through the expectation maximization (EM) algorithm. 
Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space.
%, with a clear resolution of cluster centers and without
%significant overlap of the points belonging to different clusters.
% @gui:  don't know what "clear resolution of cluster centers" adds to this?
% @jovo: commented above and included this phrase instead
Gaussian mixture models (GMM) is also very commonly used for
clustering,
however it is strongly parametric, as $k$-means which is closely related.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. 
%This enables one to cluster way more complicated datasets, 
% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
% @jovo: what I mean is not tied to any algorithm. I want to say that kernel
% methods allows one to cluster data that is not separable in Euclidean space.
% I modified the phrase, see if it's better
However, the kernel choice remains the biggest challenge
since there is no principled theory to construct a kernel for a given
dataset,
% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
and usually a kernel introduces hyperparameters that need to be carefully
chosen.
%extra parameters that need to be optimized. 
% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
% @jovo: tried to fix above, you're right in what I meant
The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

% @gui: alg 2 line 6 i think is wrong, you should do all k except j
% @gui: alg 3: we are optimizing with eq 51 the goal is to directly optimize eq 48.
% @gui: make it clear in alg 3 that the difference between it and alg 2 is the cost function, explicitly write which lines are different.
% @gui: check, post submission, whether the hardigan's results apply straightforwardly to the kernel k-means settings.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement 
%showing that there is only one possible optimization problem
%consistent with energy statistics. 
% @gui: not sure what you mean "only one possible optimization problem".  we
% can reformulate it many different ways? also, below, "this" is ambiguous.
% @jovo: I had in mind that minimizing W is equivalent to maximizing S, 
% but it's confusing at this point so just removed the phrase and adapted
%Moreover, we show that this is
leading to
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
\cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local 
solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}.
In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a family of 
default kernels, showing that
it is able to perform accurately on datasets coming very different 
distributions, contrary to $k$-means and GMM for instance.
% @gui: to me, we did not make explicit one of the most important points.   in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).  
% @jovo: added this phrase, check it
More specifically, the proposed method performs closely to $k$-means
and GMM on normally distributed data with balanced clusters, however, it
performs considerably better on  data that is not normally distributed.
It also
performs better than $k$-means and GMM in higher dimensions, even on
gaussian settings, and it performs worse than
GMM when we have highly unbalanced clusters.

% @gui: i find ToC paragraphs useless.
% @jovo: I agree and don't mind removing, however it's a tradition that 
% most referees like to see and it doesn't do any harm. I'm fine with
% whatever choice
Our work is organized as follows. In section~\ref{sec:background} we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard. We also show the equivalence to 
kernel $k$-means problem.
In Section~\ref{sec:weighted} we generalize these results to a weighted
version of energy statistics, which provides the connection with graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm based on Hartigan's method
to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

@@

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity 
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|
\end{equation}
called \emph{energy distance} \cite{Szkely2013} 
is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

@@

% @gui: let the reader decide what is important, rather than us telling them :)
% @jovo: removed that
The energy distance \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legitamate metric.

@@

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$. 
%Moreover, this can be estimated from a sample
%according to \eqref{eq:inner_data}.
% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
% @jovo: removed. I was thinking that you can always estimate the expected 
% value by computing a sample mean. but I agree, it's confusing

>>>>

\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
%\usepackage{mathtools}
%\usepackage{printlen}
%\usepackage{cite}
\usepackage[inline]{enumitem}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}

%\pretolerance=10000
%\tolerance=2000 
%\emergencystretch=10pt

@@

\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

@@

\author{Guilherme Fran\c ca}
\email{guifranca@gmail.com} 
\author{Joshua T. Vogelstein}
\email{jovo@jhu.edu}
\affiliation{Johns Hopkins University}

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a hypothesis test for equality of distributions. Energy statistics was 
further generalized to data living metric spaces of strong negative type, 
and more recently a connection with reproducing kernel Hilbert spaces (RKHS) 
was proposed. Here we consider the problem of clustering data from an 
energy statistics 
theory perspective, called energy clustering for short, by providing a 
precise mathematical formulation yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS, 
which we show to be equivalent to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and gaussian mixture models in a variety of settings.
\end{abstract}

% @gui: have other people shown a relationship between kernel k-means and 
% say, RKHS, or something else? 
% @jovo: any kernel method is defined on a RKHS. 
% For instance ref [11] is the first one to consider kernel k-means in a 
% RKHS. The problem is how do you formulate a 
% statistical problem in a RKHS, or how do you know if there is a kernel
% related to it. In the case of energy statistics this was done by [6], and
% is a nontrivial result. It's based on this work that we have the theory 
% presented in this paper

@@

% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.

% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.  k-means is certainly closely related to GMM though.

% @gui:  don't know what "clear resolution of cluster centers" adds to this?

% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?

% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise

% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?

% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.

% @gui: to me, we did not make explicit one of the most important points.   in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).

% @gui: i find ToC paragraphs useless.

% @jovo: corrected all of these points, and also re-wrote some other stuff. Please check. About ToC, I agree and wouldn't mind remove it, however it is a tradition that most referees like to see. In this case I included some important information about our work too, and it makes a transition between introduction and the remaining of the paper. If you want to remove it let me know and I will adapt.

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics. 
It provides a hypothesis test for equality of distributions which is achieved 
under minimum energy. When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution, i.e. remains finite, when they are 
equal. Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering was already proposed \cite{Kgroups}, 
which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
and references therein, for an overview of energy statistics theory and 
its applications.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
The notion of distance covariance was further generalized from Euclidean 
spaces to metric spaces of strong negative type \cite{Lyons}. Moreover, 
the missing link between energy distance based tests and kernel based tests has 
been recently resolved \cite{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence provides the link between energy distance based measures and 
RKHS based measures, commonly used in machine learning, and form the basis 
of our approach.

The clustering problem has such a long history in machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, and it is thus sensitive 
to outliers. Nevertheless, $k$-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is also 
very commonly used for clustering, however, it makes strong assumptions about 
the distribution of the data, as $k$-means which is closely related to GMM.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. Yet, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection to kernel methods, by showing that this 
QCQP is equivalent to kernel $k$-means optimization problem. The 
equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
how these connections arise from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \cite{Kgroups}, and its advantages
over Lloyd's method was already being demonstrated \cite{Telgarsky,Slonin},
but apparently it did not receive the desired attention. To the 
best of our knowledge, Hartigan's method was not previously applied together 
with kernel methods. We provide an algorithm based on Hartigan's method in 
terms of the Gram matrix of the data computed from a kernel fixed by 
energy statistics. We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to locally solve our QCQP. We show that both algorithms  have the same
time complexity.

Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering, 
or \emph{energy clustering} for short, is more accurate and robust than 
Lloyd's method. Moreover, in our experiments we put in evidence the
flexibility of energy clustering, which provides a family of default kernels, 
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, the proposed algorithms for energy clustering perform 
closely to $k$-means and GMM on normally distributed data. On the other
hand, it performs considerably better than $k$-means and GMM on  data that 
is not normally distributed. It also performs better than $k$-means 
and GMM in high dimensions, even on Gaussian settings. However, it performs 
worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
with normal distributions.

Our work is organized as follows. In section~\ref{sec:background}, we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard in general. We also show the equivalence to 
kernel $k$-means optimization problem.
In Section~\ref{sec:weighted}, we generalize these results to a weighted
version of energy statistics, providing connections to graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
problem in one dimension,
where we propose an algorithm which requires no random initialization.
In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
and propose iterative algorithms to solve the QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments showing that in several settings, energy clustering based
on Hartigan's algorithm
outperms kernel $k$-means algorithm, standard $k$-means algorithm 
and GMM with the expectation maximization algorithm. 
Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
% @jovo: removed. I was thinking that you can always estimate the expected 
% value by computing a sample mean. but I agree, it's confusing

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also said to be non-parametric, so I thought that in this context it means that it does not assume any form of the distributions.

@@

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity 
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance} \cite{Szkely2013}, 
is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

@@

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric.

@@

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$.
